---
title: "Homework 4 - Stat 495 - Fall 2025"
author: "Daizy Buluma"
date: today
format: pdf
editor: visual
---

```{r}
#| label: load-packages
#| include: false

library(mosaic)
library(tidyverse)
library(ranger)
library(gbm)
library(nnet)
library(QSARdata)
library(rsample)
```

### Homework 4 - Due Monday, Nov. 3rd by midnight to Gradescope

# Practicing Academic Integrity

If you worked with others or used resources outside of provided course material, such as Generative AI (anything besides our textbook, course materials in the repo, labs, R help menu) to help complete this assignment, please acknowledge them below using a bulleted list. Remember that according to our class framework, if you engage with Generative AI tools, the tool and date of use must be listed here, and then a complete record of the chat history must be appended to the assignment.

<!-- ~~~~~~~~~~~~~~~~ YOU MAY BEGIN EDITING BELOW THIS LINE ~~~~~~~~~~~~~~~~ -->

*I acknowledge the following individuals with whom I worked on this assignment:*

Name(s) and how they assisted you

-   

*I used the following sources to help complete this assignment:*

Source(s) and where used in your submission

-   

\newpage

```{r, include = FALSE}
library(mosaic)
options(digits = 6)
```

# PROBLEMS TO TURN IN: Add 1, Add 2, and Add 3

The first two problems use the bbb2 data set from the QSARdata package for Chapter 17 and 18 material. The final problem covers concepts from Chapter 15.

For the applied problems, the response variable of interest is bbb2_Class, which can be found in the bbb2_Outcome data set. We have joined the outcome variable to the QuickProp data set that we want to focus on below. You can read the associated help file in R to learn more about the data set. You saw part of this data set on the exam.

```{r}
data(bbb2)
#?bbb2 # for variable reference and information
mybbb2 <- left_join(bbb2_QuickProp, bbb2_Outcome) %>% dplyr::select(- Molecule)
tally(~ bbb2_Class, data = mybbb2)
```

Our goal for the applied problems below is to use the recent methods from class (Chapter 17 and 18) to predict the response variable, whether each compound "crosses" the blood-brain barrier or "does not" cross. You should use the *mybbb2* data set going forward. Note that there are 51 variables at the moment, and the last variable "Class" is the target, but if you open the data set to view, it will only show 50 variables by default. Class is there, but you have to use the arrows to see it.

```{r}
# We loaded a lot of data sets we don't need anymore
# remove them to clean up your workspace
remove(bbb2_AtomPair, bbb2_Daylight_FP, bbb2_Dragon, bbb2_Lcalc, bbb2_moe2D,
       bbb2_moe2D_FP, bbb2_moe3D, bbb2_Outcome, bbb2_PipelinePilot_FP,
       bbb2_QuickProp)
```

To avoid issues with reproducibility, you should set a seed in EACH chunk below where you do a random process, whether that is setting up the train/test split or fitting a model that has some random process involved.

Note that because some of these methods can take a long time to tune, we will NOT be tuning for them. Instead, you will run them using their original packages/functions *ranger*, *gbm*, and *nnet* with the inputs specified. This is to introduce you to those different inputs and what they are for/what they do. If you end up applying these models as some part of the final project, be sure you tune the models appropriately! You should use the help function for information about the functions if you are having issues running the models as a starting point.

\newpage

## Additional Problem 1

Your task for this problem is to do some setup, then fit the models described in parts c, d and e, and then compare them in part f. Use all available predictor variables (except what is removed in part a), with no re-expressions. The same training/test split will be used in Additional Problem 2 as well (i.e. you only make this once).

> part a. One variable in the data set, QikProp\_.amidine causes issues with some of these methods. We will remove it here, but can you see why it is problematic? Explain why this variable is not very useful for this analysis.

SOLUTION: The variable `QikProp_.amidine` has almost no variation. Out of 80 observations, 79 are 0 (98.75%) and only 1 is 1 (1.25%). A predictor with essentially only one level provides no useful information for distinguishing between outcomes (near zero-variance variables). Therefore, we exclude this variable before modeling because it will not contribute to predictive performance and may cause issues for some methods.

```{r}
mybbb2 |> 
  count(QikProp_.amidine) |> 
  mutate(percent = n / sum(n) * 100)
```

```{r}
# run once you are ready to remove the variable to proceed
# this variable is not used anywhere below, so overwrite the data set

mybbb2 <- dplyr::select(mybbb2, -QikProp_.amidine)
```

> part b. Create an appropriate training/test split from mybbb2 with a ratio of 70/30 to use throughout the problems. As always, be sure your split is reproducible.

SOLUTION:

```{r}
set.seed(008)

split <- initial_split(mybbb2, prop = 0.70, strata = "Class")
cross_train <- training(split)
cross_test <- testing(split)

tally(~ Class, data = cross_train, format = "count")
```

> part c. Create an appropriate model to predict Class with the training set using bagging with 1000 trees and otherwise with default settings for tuning parameter values.

SOLUTION:

```{r}
set.seed(008)
bag_cross <- ranger(
  formula = Class ~ ., 
  data = cross_train, 
  mtry = 49, #number of predictors
  num.trees = 1000
)
#summary(bag_cross)
bag_cross
```

> part d. Create an appropriate model to predict Class with the training set using a random forest with 1000 trees and otherwise with default settings for tuning parameter values.

SOLUTION:

```{r}
set.seed(008)
rf_cross <- ranger(
  formula = Class ~ ., 
  data = cross_train, 
  num.trees = 1000
)
rf_cross
```

> part e. Create an appropriate model to predict Class with the training set using boosting with 500 trees and otherwise with default settings for tuning parameter values.

SOLUTION:

```{r}
cross_train$Class_num <- ifelse(cross_train$Class == "Crosses", 1, 0)
cross_test$Class_num <- ifelse(cross_test$Class == "Crosses", 1, 0)

set.seed(008)
boost_cross <- gbm(
  formula = Class_num ~ ., 
  data = cross_train |> dplyr::select(-Class), 
  n.trees = 500,
  distribution = "bernoulli") 
boost_cross
```

> part f. Parts c, d, and e only required you to fit models. Now we want to compare their performance. Use an appropriate measure to compare the three models in terms of their model performance based on the test set. Be sure your choice of measure is clear. Summarize your findings. Then discuss which model you would choose for predicting Class. Explain your choice. (You can use additional assessment items, but must have the test set comparison.)

SOLUTION:

I used the estimated True Error Rate (TER) to compare the performance of the three models. The bagging and boosting models both achieved about 84% test accuracy, giving an estimated TER of about 0.16. The random forest model had slightly lower test accuracy (around 80%), leading to an estimated TER of about 0.20. Based on this test set comparison, bagging and boosting appear to perform the best.

Although bagging and boosting had the same estimated TER, I would choose the bagging model for predicting *Class* because it produced strong predictive performance and was simpler to implement and evaluate. Also, boosting generally requires more tuning than bagging and can easily overfit if the hyperparameters are not carefully selected.

```{r}
#for bagging - 84%
bag_test_pred <- predict(bag_cross, cross_test)$predictions
confusionMatrix(bag_test_pred , cross_test$Class)

#for random forest - 80%
rf_test_pred <- predict(rf_cross, cross_test)$predictions
confusionMatrix(rf_test_pred , cross_test$Class)

#for boosting
boost_data_test <- cross_test %>% dplyr::select(-Class)
boost_test_pred <- predict(boost_cross, 
                           newdata = boost_data_test, 
                           n.trees = 500,
                           type = "response")

# Convert predictions to factor with explicit mapping
boost_class <- factor(ifelse(boost_test_pred > 0.5, 1, 0),
                      levels = c(0,1),
                      labels = c("DoesNot", "Crosses"))

# Convert true test labels using the SAME mapping
boost_data_test$Class_num <- factor(boost_data_test$Class_num,
                                    levels = c(0,1),
                                    labels = c("DoesNot", "Crosses"))



# Confusion matrix
confusionMatrix(boost_class, boost_data_test$Class_num, positive = "Crosses")
```

```{r}
cross_train <- cross_train %>% dplyr::select(-Class_num)
cross_test  <- cross_test %>% dplyr::select(-Class_num)
```

\newpage

## Additional Problem 2

Your task for this problem is to fit the models described in parts a, b, and c, and then compare them in part d. Use all available predictor variables, with no re-expressions. Use the same training/test data as above. Finally, in part e, you'll try to do better with a model of your own.

(Remember, these tuning parameter values were arbitrary choices I made so you can see the inputs and such to the models. They are not tuned for optimal performance!)

> part a. Create an appropriate model to predict Class with the training set using a neural net with a single hidden layer of 12 nodes, and otherwise with default settings for tuning parameter values.

SOLUTION:

```{r}
set.seed(008)
nnet_cross <- nnet(Class ~ ., cross_train, size = 12)
```

> part b. Create an appropriate model to predict Class with the training set using a neural net with a single hidden layer of 12 nodes, and a decay parameter of 6e-3, and otherwise with default settings for tuning parameter values.

SOLUTION:

```{r}
set.seed(008)
nnet_cross2 <- nnet(Class ~ ., cross_train, size = 12, decay = 6e-3)
```

> part c. Create an appropriate model to predict Class with the training set using a neural net with a single hidden layer of 12 nodes, a decay parameter of 6e-3, and a value for maxit that allows for convergence, and otherwise with default settings for tuning parameter values. (Ask if you aren't sure what the convergence idea is about!)

SOLUTION:

```{r}
set.seed(008)
nnet_cross3 <- nnet(
  Class ~ ., 
  data = cross_train, 
  size = 12, 
  decay = 6e-3,
  maxit = 630)
```

> part d. Parts a, b, and c only required you to fit models. Now we want to compare their performance. Use an appropriate measure to compare the three models in terms of their model performance based on the test set. Be sure your choice of measure is clear. Summarize your findings. Then discuss which model you would choose for predicting Class. Explain your choice.

SOLUTION:

I used test set accuracy as the performance measure. Based on the confusion matrices, the `nnet_cross` achieved a test accuracy of about 0.80, which corresponds to an estimated TER of 0.20. `nnet_cross2` achieved a test accuracy of about 0.76 and an estimated TER of 0.24. Finally, `nnet_cross3`, also achieved a test accuracy of 0.80 and therefore the same estimated TER of 0.20.

`nnet_cross`and `nnet_cross3` both converge and achieve the same accuracy on the test set. However `nnet_cross3` includes weight decay, which shrinks weights and helps guard against overfitting and typically yields more stable performance across splits. I would therefore choose `nnet_cross3.`

```{r}
# nnet_cross
pred1 <- predict(nnet_cross, newdata = cross_test, type = "class")
pred1 <- factor(pred1, levels = levels(cross_test$Class))
confusionMatrix(pred1, cross_test$Class)

# nnet_cross2
pred2 <- predict(nnet_cross2, newdata = cross_test, type = "class")
pred2 <- factor(pred2, levels = levels(cross_test$Class))
confusionMatrix(pred2, cross_test$Class)

# nnet_cross3
pred3 <- predict(nnet_cross3, newdata = cross_test, type = "class")
pred3 <- factor(pred3, levels = levels(cross_test$Class))
confusionMatrix(pred3, cross_test$Class)
```

> part e. Fit a new neural net model with parameters chosen by you (you can even explore options we didn't use above), but must set the size of the hidden layer and ensure your model reaches convergence. What parameter values did you use? Does your model do better than your final model in part d.?

SOLUTION: No, it achieves the same test accuracy.

```{r}
set.seed(008)
nnet_cross4 <- nnet(
  Class ~ ., 
  data = cross_train, 
  size = 12,
  decay = 0.001,   
  maxit = 1000) 


pred4 <- predict(nnet_cross4, newdata = cross_test, type = "class")

# Match factor levels
pred4 <- factor(pred4, levels = levels(cross_test$Class))

# Accuracy / confusion matrix
conf_e <- confusionMatrix(pred4, cross_test$Class, positive = "Crosses")
conf_e
```

\newpage

## Additional Problem 3

In a few sentences, explain to someone currently taking STAT 135 (so their level of background knowledge is intro stats, just having seen hypothesis testing for the first time), what p-value adjustment is all about and why it's an important concept to understand.

SOLUTION:

As you know, when performing a hypothesis test, we usually calculate a p-value to help us decide whether our result is statistically significant or whether occurred just by random chance. We compare this p-value against a threshold, alpha, which represents the probability of making a Type I error (incorrectly concluding there's an effect when there really isn't one). Now suppose we set $\alpha$ = 0.05 and test 100 different variables to see if any are related to an outcome. Even if none of them are truly related, we'd still expect about five of them to show a “significant” result just by chance because 0.05 means there’s a 5% chance of a false positive for each test.

When you run multiple tests, each one carries its own 5% risk. Those risks start to add up and the overall chance of making at least one false positive across all tests becomes much higher than 5%. This where p-value adjustment comes in. It helps bring that overall risk back down to a reasonable level. This is done by making the criteria for significance stricter when performing multiple comparisons. There are several adjustment methods, but they all aim to reduce the likelihood of mistakenly claiming a discovery when the result is really just noise

Understanding p-value adjustment is important because studies often involve multiple hypothesis tests at once. Without it, we might believe we’ve found something meaningful when, in reality, the result occurred by random chance. The adjustment gives you more confidence that the results you call “significant” are actually meaningful.
