---
title: "Data Analysis Project"
author: "Daizy Buluma"
date: today
format: pdf
editor: visual
---

```{r}
#| label: load-packages
#| include: false

library(mosaic)
library(tidyverse)
library(lubridate)
library(GGally)
library(viridis)
library(leaflet)
library(caret)
library(rsample)
library(rpart)
library(MASS)
library(dplyr)
library(car)
library(rpart)
library(rpart.plot)
```

### Project - Due Friday, Oct. 24th by midnight to Gradescope

# Practicing Academic Integrity

If you worked with others or used resources outside of provided course material, such as Generative AI (anything besides our textbook, course materials in the repo, labs, R help menu) to help complete this assignment, please acknowledge them below using a bulleted list. Remember that according to our class framework, if you engage with Generative AI tools, the tool and date of use must be listed here, and then a complete record of the chat history must be appended to the assignment.

<!-- ~~~~~~~~~~~~~~~~ YOU MAY BEGIN EDITING BELOW THIS LINE ~~~~~~~~~~~~~~~~ -->

*I acknowledge the following individuals with whom I worked on this assignment:*

Name(s) and how they assisted you

-   

*I used the following sources to help complete this assignment:*

Source(s) and where used in your submission

-   

\newpage

## Instructions for your submission

Your submission should include the following sections (along with all code necessary to reproduce your results):

-   Introduction and Exploratory Data Analysis - could be two separate sections
-   Your final GLM and relevant details
-   Your final Tree and relevant details (can be before or after your GLM)
-   Your Model Comparison (can be woven in sections above)
-   Your Conclusion

Descriptions of the purpose for each section follow. Do not use this as a set of check boxes for the assignment. Instead, the idea here is to explain why you'd write each section, and you can work out what needs to be in each in order to fulfill that purpose.

-   Introduction - The real estate developer has interests, but will you be able to address them? How do you understand the tasks you are presented with? It's important to state what you will be doing in the analysis, so the reader can make sure their understanding lines up with what you will be doing. The reader also needs an introduction to your data, either in this section, or below. They need enough detail to be able to determine if your actions later in the analysis are reasonable.

-   Exploratory Data Analysis - You are the analyst here. That means you can use variable transformations, subsets of observations, and employ other modeling/analytic practices (e.g. training/testing data sets) at your discretion to aid in your analysis, so long as you explain your rationale for them. The reader needs to know what you found when looking at the data and what you decided to do about it, because this has impacts on the rest of your analysis. Remember our lessons from the first classes - be sure you look at the data! Here's your chance to share what you found based on how it impacts your analysis.

-   GLM and Tree sections - These are new techniques. By having each in their own section, you can focus on your explanation for them one at a time. They can be in any order. The developer does not know what these are - you need to explain them at least enough they can follow your discussion. You have options that you need to pick for the different techniques (various tuning parameters) - for example, are you using a classification or a regression tree? What tree stopping options are being used? What input variables are you using and why? What type of GLM are you using? How will you be measuring performance? Be sure you discuss what options you are selecting and what made you choose those values. Helping explain this shows your understanding of the techniques. (Adding this level of detail is a feature of this assignment that makes it a little different from the example data analysis articles you looked at.)

You also want these sections to show off your ability to build models. It is not appropriate to only fit kitchen sink models or accept all default settings without exploring to see if that is really what is best for the task at hand. Your write-up should make it clear what you explored, but you don't have to show every model you considered. You can do a lot in sentences discussing your process. How you got to your final model should be clear.

Finally, remember to check out your "best" model for each technique. Did you check reasonable diagnostics? Does the model make sense? Are variables behaving as you expect? If you focus solely on model performance, that's missing the point. For example, you might find a model has very little error because a variant of the response was accidentally included as a predictor. You are responsible for checking over the model before reporting it. Your write-up should convey that you've done this (the reader can't read your mind to know you did it).

-   Model Comparison - There are several ways to compare the models - performance, interpretability, variables involved, etc. Remember you are trying to address the real estate developer's questions of interest. You may do that with a focus on just one model, or maybe you found a way to combine results from a "best" GLM and a "best" tree. The idea for this section is that you need to convey the process used to pick a final model(s) to use to answer those questions, with justification for your choices. This section can be woven in the tree/GLM sections so it may not be a separate section.

-   Conclusion - This section should be a stand alone summary of your analysis. It is where you get to state your final model and a quick summary of the process used to get there. You also need to address the developer's questions (if not done before), and this is your last place to do that! Remember to flush out the details here. If your final sentence is "Model 5 is the model I choose and it answers your questions", does it? Are you doing your job as the analyst to leave things at that? For that matter, what is Model 5?

-   Printing Trees - It is fine to print your trees out to separate .pdfs if you want to for some reason (just be sure to include them in your submission!). Please do show your code though (so no echo = FALSE) and remember that your work should be reproducible.

-   Length - Your submission should be as long as it needs to be, but not unnecessarily long. Plots, etc. are encouraged to help illustrate points, but don't include output that isn't needed. For example, a histogram or densityplot and favstats for each variable is something good to explore, but you shouldn't put that in your submission for every variable. You might need plots to point out outliers you are removing, or maybe you show one plot and then say others were similar. I encourage you to do any additional work in a separate .Qmd from your submission.

-   Audience - For purposes of this submission, remember that the audience is the real estate developer, so you'll probably need to include some explanations that you'd leave out of another assignment. For example, the developer isn't going to know what a GLM is, or what type you are using, or why you'd use it. You should think about where that information should go, and do your best to explain what you need in order to report your findings. Similarly, assume that you found this data to assist the developer and they need basic details about it to understand the variables. They didn't hand it over to you.

You may remove these instructions if desired for your submission, but must leave the page on Academic Integrity above. The template with sections for the report is below.

\newpage

# Introduction

\newpage

# Exploratory Data Analysis

```{r}
#| include: false
# Can be included with the introduction

kchouse <- read.csv("https://awagaman.people.amherst.edu/stat495/kc_house_data.csv",
                    header = T)
```

I started by examining the structure and basic characteristics of the dataset to understand its scale and completeness before modeling. The dataset contains 21,597 records and 21 variables that describe property attributes such as living area, lot size, number of rooms, and quality ratings, along with geographic and temporal identifiers.

```{r}
glimpse(kchouse)
```

The dataset has no missing values.

```{r}
sum(is.na(kchouse))
```

```{r}
#make it binary(?)
summary(kchouse$view)
unique(kchouse$view)
```

```{r}
#make it binary(?)
summary(kchouse$waterfront)
unique(kchouse$waterfront)
```

```{r}
summary(kchouse)
filter(kchouse, bedrooms == 33 | price == 7700000)
#drop this record?
```

Some key observations that stand out are:

1.  Sale prices range from about \$78,000 to \$7.7 million, with a median of \$450,000 and a mean of \$540,000. Roughly half of all homes fall above the \$500,000 threshold, meaning the “high-price” and “not-high-price” classes will be reasonably balanced for classification.
2.  Most homes have 3 bedrooms and 2 bathrooms, which is typical of most family houses. However, there are some striking outliers such as a property with 33 bedrooms and a lot over 1.6 million sqft (\~ 38 acres). These extreme values will be monitored as analysis continues.
3.  The properties were built between 1900 and 2015. Notably very few have been renovated, which suggests that whether a house has been renovated could hold predictive value. To capture potential effects of property upgrades, we created a binary variable `has_renovated` indicating whether a home had any recorded renovation (`yr_renovated > 0`). Only 4% of the properties have been renovated and the boxplot shows that renovated homes tend to command slightly higher prices on average, suggesting that renovation status could be a useful predictor of high-price classification

```{r}
kchouse <- kchouse |> 
  mutate(has_renovated = factor(if_else(yr_renovated > 0, "Renovated", "Not Renovated")))

# Quick check of how many homes were renovated
count(kchouse, has_renovated) |> 
  mutate(percent = n / sum(n) * 100)

ggplot(kchouse, aes(x = has_renovated, y = price)) +
  geom_boxplot() +
  labs(title = "Price Comparison: Renovated vs. Not Renovated",
       x = "", y = "Price ($)")


```

4.  The `date` variable was read as text rather than a proper date format. After conversion, it reveals sales primarily between May 2014 and May 2015.

    ```{r}
     kchouse <- kchouse |> 
       mutate(date = mdy(date)) 
       summary(kchouse$date)
    ```

Next, the target variable `highprice` was created to distinguish homes selling above versus below \$500,000. Out of 21,597 home sales, about 42% sold for more than \$500,000, while 58% sold for less. This indicates that most homes in King County were still below the half-million threshold during the 2014–2015 period, but the high-price segment was substantial. Although, the target variable is not severely imbalanced enough to cause modeling issues, we'll account for this difference during model evaluation.

```{r}
kchouse <- kchouse |> 
           mutate(high_price = factor( if_else(price > 500000, "Above $500K", "Below $500K")))

count(kchouse, high_price) |> 
  mutate(percent = n / sum(n) * 100)               
```

```{r}
ggplot(kchouse, aes(x = high_price, fill = high_price)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Homes Above vs. Below $500K",
    x = "",
    y = "Count",
    fill = "Price Category"
  )
```

```{r}
mean_lat <- mean(kchouse$lat)
mean_long <- mean(kchouse$long)

leaflet(kchouse) |> 
  addTiles() |> 
  setView(lng = mean_long, lat = mean_lat, zoom = 10) |> 
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    color = ~ifelse(price > 500000, "red", "blue"),
    radius = 3,
    opacity = 0.8,
    fillOpacity = 0.6
  )
```

The spatial visualization reveals a distinct clustering of high-priced homes (red) along Seattle’s waterfronts and around Lake Washington, while lower-priced homes (blue) dominate inland and southern areas such as Renton and Kent. This geographic pattern aligns with the region’s economic landscape urban neighborhoods near downtown Seattle and the Eastside (e.g., Bellevue and Kirkland) host major technology hubs and high-income employment centers, driving up property demand and prices. To capture the influence of water proximity observed on the map, I created a categorical indicator variable, `waterfront_flag`, distinguishing between “Waterfront” and “Not Waterfront” properties. The variable is stored as a factor to ensure it is correctly treated as a categorical predictor in subsequent modeling.

```{r}
kchouse <- kchouse |> 
  mutate(waterfront_flag = factor(if_else(waterfront > 0, "Waterfront", "Not Waterfront")))


ggplot(kchouse, aes(x = waterfront_flag , y = price)) +
  geom_boxplot() +
  labs(title = "Price Comparison: Waterfront vs. Not Waterfront",
       x = "", y = "Price ($)")
```

The boxplot comparing waterfront and non-waterfront properties shows a substantial price gap between the two groups. Waterfront homes have a notably higher median sale price and a broader interquartile range, with many properties exceeding \$2 million. In contrast, most non-waterfront homes cluster below \$1 million, with a few extreme luxury outliers. This reinforces the insight from the spatial map. The clear separation between the two distributions further justifies including the waterfront_flag variable as a categorical predictor in the modeling phase.

```{r}
names(kchouse)
```

Finally, I visualized pairwise relationships among the most relevant numeric features in the dataset to better understand how the main predictors relate to each other and to the target variable. This helps identify strong linear associations and potential multicollinearity before moving into modeling phase. Only variables that explain home prices based on property size and quality are included. The following variables are excluded because of the provided reasons: \* `id` - a unique identifier with no analytical meaning. \* `date` - represents sale date. \* `zipcode`, `lat`, and `long` - spatial variables which have been analyzed above. \* `sqft_living15` and `sqft_lot15` - 2015 measurements highly correlated (\>0.7) with their original counterparts, excluded to avoid redundancy and multicollinearity.

```{r}
cor(kchouse$sqft_living, kchouse$sqft_living15)
cor(kchouse$sqft_lot, kchouse$sqft_lot15)
```

-   `view` - excluded due to lack of measurement context. It is uncertain whether the “view” rating reflects a physical inspection, an online listing impression. Including it without clarity could introduce noise rather than meaningful information.
-   `has_renovated`, `high_price`, and `waterfront_flag` - are categorical indicators which have already been explored above.

```{r}
num_vars <- kchouse |> 
  dplyr::select(price, sqft_living, sqft_lot, bedrooms, bathrooms, floors,
         condition, grade, sqft_above, sqft_basement, yr_built)

# Pairwise plot matrix colored by high_price
ggpairs(
  kchouse,
  columns = c("price", "sqft_living", "bathrooms", "bedrooms", 
              "grade", "sqft_lot", "floors", "condition", "yr_built"),
  aes(color = high_price, alpha = 0.5),
  progress = FALSE
)
```

The correlation matrix reveals several strong relationships between predictors and the target variable. `sqft_living` and `grade` have the strongest correlations with `price` ($\approx$ 0.70 and 0.67), indicating that larger and higher-quality homes sell for significantly more. `yr_built`, `condition`, and `sqft_lot` have relatively low correlations with `price`, implying limited direct influence on value. The pairwise plots also reveal several instances of potential multicollinearity among `sqft_living`, `bathrooms`, and `bedrooms`. This is expected as all variables provide overlapping information on the home size. Variables showing excessive correlation (e.g., VIF \> 5) may be dropped in the next phase to improve model stability and interpretability.

\newpage

## Training and Testing Setup

The variables (`id`, `date`, `price`, `zipcode`, `lat`, `long`, `view`, `yr_renovated`, `waterfront`) were removed due to reasons explained in the previous section. 

```{r}
kchouse_clean <- kchouse |> 
  dplyr::select(-id, -date, -lat, -long, -zipcode, -price, -yr_renovated, -waterfront, -view, -sqft_living15, -sqft_lot15)

set.seed(22)
# Set up data set train/test split

split <- initial_split(kchouse_clean, prop = 0.70, strata = "high_price")

house_train <- training(split)
house_test <- testing(split)

myControl <- trainControl(method = "repeatedcv", 
                          number = 10, 
                          repeats = 10,
                          classProbs = TRUE)
```

```{r}
tally(~ high_price, data = house_train, format = "count")
```

Baseline misclassification rate computation:

```{r}
6335/(6335+8782)
```
If we classified all properties in the training set to below \$500K (the majority class), our error rate on training would be 41.9%. Our models need to do better than this.

# GLM and Results

After splitting the dataset into training and testing sets, the next step was to fit a logistic regression model to predict whether a house is classified as high-priced (above $500,000) or not. Logistic regression is appropriate here because the response variable, `high_price`, is binary (0 = low-priced, 1 = high-priced).

Before fitting the model, we checked for multicollinearity, a situation where predictors are highly correlated, which can inflate standard errors and make coefficient estimates unstable.

We used the `alias()` function to detect perfect linear relationships among predictors and found that `sqft_living`, `sqft_above`, and `sqft_basement` were perfectly correlated (`sqft_living = sqft_above + sqft_basement`).
To prevent aliasing, only `sqft_living` was retained, as it represents the total interior space and captures the same information.
After cleaning the dataset and removing redundant variables (`id`, `date`, `lat`, `long`, `zipcode`, `view`, `price`, and the aliased columns), the model predictors included: bedrooms, bathrooms, sqft_living, sqft_lot, floors, condition, grade, yr_built, has_renovated, waterfront_flag'

```{r}
pre_glm_full <- glm(
  high_price ~ .,
  data = house_train,
  family = binomial
)
alias(pre_glm_full)
```

```{r}
glm_full <- glm(
  high_price ~ bedrooms + bathrooms + sqft_living + sqft_lot +
    floors + condition + grade + yr_built + has_renovated + waterfront_flag,
  data = house_train,
  family = binomial
)

# Check for multicollinearity
summary(glm_full)
vif_values <- vif(glm_full)
vif_values
```
The Variance Inflation Factor (VIF) quantifies how much variance in a coefficient is inflated due to multicollinearity.
A VIF value above 5 typically signals problematic correlation. In this case, all predictors had VIF values below 5, suggesting that multicollinearity was not a concern after removing redundant variables.

To simplify the model while maintaining predictive power, stepwise selection using AIC was applied.
The AIC (Akaike Information Criterion) balances model fit and complexity, a lower AIC is better and preferred.

```{r}
# Stepwise model selection using AIC
glm_step <- stepAIC(glm_full, direction = "both", trace = FALSE)
glm_int <- glm(
  high_price ~ bedrooms + bathrooms + sqft_living + sqft_lot +
    floors + grade + yr_built + has_renovated * condition  + waterfront_flag,
  data = house_train,
  family = binomial
)
summary(glm_int)
anova(glm_step, glm_int)
# Extract and display best model
best_model <- formula(glm_step)
summary(glm_step)
```
The stepwise AIC procedure automatically removed `has_renovated` as it did not significantly contribute to predicting `high_price`. The resulting model retained predictors that are the strongest indicators of a house being high-priced.

To assess generalization performance, we used 10-fold cross-validation repeated 5 times to ensure that performance metrics are not dependent on a single random data split.
The same training/test and CV framework will later be used for the decision tree model, ensuring fair comparison.
```{r}
house_train$high_price <- factor(house_train$high_price,
                                 levels = c("Below $500K", "Above $500K"),
                                 labels = c("Below500K", "Above500K"))

house_test$high_price <- factor(house_test$high_price,
                                levels = c("Below $500K", "Above $500K"),
                                labels = c("Below500K", "Above500K"))
```

```{r}
#ask if this is redundant work
set.seed(22)
glm_cv <- train(
  best_model,
  data = house_train,
  method = "glm",
  family = "binomial",
  trControl = myControl
)

glm_cv
```
For this model, the mean accuracy was approximately 81%, indicating strong predictive ability relative to the baseline classification rate of 41.9%.


#more tests to be done to find better model e.g interactions

\newpage

# Tree and Results

```{r, include = FALSE}
# Can be before or after your GLM
# You should pick what you want to present first
tree_grid <- expand.grid( cp =c(0.01, 0.05, 0.10, 0.2, 0.3, 0.4))
set.seed(22)

tree_model <- train(
  high_price ~ bedrooms + bathrooms + sqft_living + sqft_lot +
    floors + condition + grade + yr_built + has_renovated + waterfront_flag,
  data = house_train,
  method = "rpart",
  trControl = myControl,
  tuneGrid = tree_grid
)

tree_model

```

```{r}
ggplot(tree_model)
```
```{r}
kchouse.rpart <- rpart(high_price ~ bedrooms + bathrooms + sqft_living + sqft_lot +
    floors + condition + grade + yr_built + has_renovated + waterfront_flag, 
                     data = house_train,
                     control = rpart.control(cp = 0.005))
printcp(kchouse.rpart)
rpart.plot(kchouse.rpart)
```
```{r}
# Predictions on test set
confusionMatrix(predict(kchouse.rpart, newdata = house_test, type = "class"), house_test$high_price)

```
About 79.2% of all test samples were classified correctly. This is much higher than the baseline “No Information Rate” (which is 0.5809), meaning the model performs much better than guessing the majority class. The model achieved slightly lower than the logistic regression, but provided better interpretability through explicit decision rules.
\newpage

# Model Comparison

```{r, include = FALSE}
# Can be woven in sections above; doesn't have to be it's own section
```

\newpage

# Conclusion
